---
title: "Alannah's R Project Part 2"
output: html_notebook
---

```{r} 
# Reading in the community data and installing picante.
install.packages("picante", dependencies = TRUE)

comm <- read.csv("grassland_data/grassland.community.csv", header = TRUE, row.names = 1)

class(comm)
dim(comm)
rownames(comm)
head(colnames(comm))
apply(comm, 1, sum)
comm <- decostand(comm, method = "total")
apply(comm, 1, sum)
comm[1:5, 1:5]
```

Reading in the trait data.
```{r}
traits <- read.csv("grassland_data/species.traits.csv", header = TRUE, row.names = 1)

head(traits)

pairs(traits)

traits <- log10(traits)
pairs(traits)
```

Reading in the metadata
```{r}
metadata <- read.csv("grassland_data/plot.metadata.csv", header = TRUE, row.names = 1)
head(metadata)
```

Reading in the phylogeny data and taking a closer look.
```{r}
#installing packages required for phylogeny visualization.
install.packages("ape")
library(ape)

phy <- read.tree("grassland_data/grassland.phylogeny.newick")
class(phy)
names(phy)
phy$tip.label[1:5]
Ntip(phy)
plot(phy, cex =0.5)
```

Cleaning and matching datasets 
```{r}
library(picante)
ls()
combined <- match.phylo.comm(phy, comm)
phy <- combined$phy
comm <- combined$comm
combined <- match.phylo.data(phy, traits)
phy <- combined$phy
traits <- combined$data
all.equal(rownames(comm), rownames(metadata))
metadata <- metadata[rownames(comm), ]
```

Visualizing and summarizing biodiversity data
```{r}
boxplot(specnumber(comm) ~ metadata$rel.moisture, ylab = "# of species")
t.test(specnumber(comm) ~ metadata$habitat)
plot(specaccum(comm), xlab = "# of samples", ylab = "# of species")

```

Multivariate community analysis 

Ordination is "ordaining" data to a matrix that will allow a closer look at community composition. Ordiplots can be used in various ways to assess how well the data is relating the community and its species. The red crosses indicate the species, so we can appreciate the different clusters formed. The addition of the environmental and trait data also adds to these clustering patterns and easing visualization.

```{r}
# A) Hierarchical clustering
comm.bc.dist <- vegdist(comm, method = "bray")
comm.bc.clust <- hclust(comm.bc.dist, method = "average")
plot(comm.bc.clust, ylab = "Bray-Curtis dissimilarity")

#Ordination
comm.bc.mds <- metaMDS(comm, dist = "bray")
stressplot(comm.bc.mds)
ordiplot(comm.bc.mds, display = "sites", type = "text")
ordipointlabel(comm.bc.mds)

mds.fig <- ordiplot(comm.bc.mds, type = "none")
points(mds.fig, "sites", pch = 19, col = "green", select = metadata$rel.moisture)
ordiellipse(comm.bc.mds, metadata$rel.moisture, conf = 0.95, label = TRUE)
ordicluster(comm.bc.mds, comm.bc.clust, col = "gray")

ordisurf(comm.bc.mds, comm[, "Sphaeralcea_coccinea"], bubble = TRUE, main = "Sphaeralcea coccinea abundance", 
    cex = 3)

#Adding environmental and trait data to ordination.
ordiplot(comm.bc.mds)
plot(envfit(comm.bc.mds, metadata[, 3:6]))

```

Trait evolution
```{r}
# Phylogenetic signal
apply(traits, 2, Kcalc, phy)
multiPhylosignal(traits, multi2di(phy))

#Visualizing trait evolution
plot(phy, direction = "up", show.tip.label = FALSE, show.node.label = TRUE, 
    cex = 0.7)
tiplabels(pch = 19, col = "black", cex = 3 * (traits[, "LeafArea"]/max(traits[, 
    "LeafArea"])))

#Phylogenetic analysis of trait relationships
root.gls <- gls(RootTissueDens ~ SRL, data = traits)
anova(root.gls)
root.pgls <- gls(RootTissueDens ~ SRL, correlation = corBrownian(value = 1, 
    phy), data = traits)
anova(root.pgls)
plot(RootTissueDens ~ SRL, data = traits, xlab = "SRL (specific root length)", 
    ylab = "Root tissue density")
abline(coef(root.gls), lwd = 2, col = "black")
abline(coef(root.pgls), lwd = 2, col = "red")
legend("bottomleft", legend = c("GLS fit", "Phylogenetic GLS fit"), lwd = 2, 
    col = c("black", "red"))

```

Phylogenetic and trait diversity

Cophenetic is a type of distance determination for the aforementioned hierarchial clustering, only this time, looking at dissimilarity amongst individual clusters.

Here, we assigned that calculation to phy.dist, then plotted against relative moisture. These are visualized as a scatterplot with an abline to look at trend.

```{r}
# Phylogenetic diversity
comm.pd <- pd(comm, phy)
head(comm.pd)
boxplot(comm.pd$PD ~ metadata$rel.moisture, xlab = "Habitat", ylab = "Faith's PD")
t.test(comm.pd$PD, metadata$rel.moisture)
plot(comm.pd$PD ~ comm.pd$SR, xlab = "Species richness", ylab = "Faith's PD")

#MPD, MNTD, SESMPD and SESMNTD
phy.dist <- cophenetic(phy)
comm.sesmpd <- ses.mpd(comm, phy.dist, null.model = "richness", abundance.weighted = FALSE, 
    runs = 999)
head(comm.sesmpd)

comm.sesmntd <- ses.mntd(comm, phy.dist, null.model = "richness", abundance.weighted = FALSE, 
    runs = 999)
head(comm.sesmntd)

plot(comm.sesmpd$mpd.obs.z ~ metadata$rel.moisture, xlab = "Relative Mositure", ylab = "SES(MPD)")
abline(h = 0, col = "gray")

t.test(comm.sesmpd$mpd.obs.z, metadata$rel.moisture)

plot(comm.sesmntd$mntd.obs.z ~ metadata$rel.moisture, xlab = "Relative Moisture", ylab = "SES(MNTD)")
abline(h = 0, col = "gray")
t.test(comm.sesmntd$mntd.obs.z, metadata$rel.moisture)

plot(phy, show.tip.label = FALSE, main = "Relative Moisture")
tiplabels(tip = which(phy$tip.label %in% colnames(comm)[comm["relative moisture", ] > 
    0]), pch = 19)

plot(phy, show.tip.label = FALSE, main = "Relative moisture")
tiplabels(tip = which(phy$tip.label %in% colnames(comm)[comm["Relative moisture", ] > 
    0]), pch = 19)

#Trait diversity
trait.dist <- as.matrix(dist(scale(traits), method = "euclidean"))
comm.sesmpd.traits <- ses.mpd(comm, trait.dist, null.model = "richness", abundance.weighted = FALSE, 
    runs = 999)

plot(comm.sesmpd.traits$mpd.obs.z ~ metadata$rel.moisture, xlab = "Relative moisture", ylab = "Trait SES(MPD)")
abline(h = 0, col = "gray")

```

Phylogenetic beta-diversity

The mantel analysis observes two different distance matrices, (a pair of them, since this is beta-diversity assessment) to evaluate the phylogenetic and trait relationships for this community.It outputs the r-value and the p-value. 

```{r}
comm.mntd.dist <- comdistnt(comm, phy.dist, abundance.weighted = TRUE)
comm.mntd.traits.dist <- comdistnt(comm, trait.dist, abundance.weighted = TRUE)
mantel(comm.bc.dist, comm.mntd.dist)
mantel(comm.bc.dist, comm.mntd.traits.dist)
```

Phylogeny/trait-based ordinations

NMDS (non-metric multidimensional scaling). Using ordination, this function can help determine community composition while showing pairwise dissimilarity, even with the many different data. This is done by using ranking of the data, which shows the grouping of species, while those more "dissimilar" are further from one another, based on both phylogeny and traits. 

From the results here, we can visualize the different habitat types and relation to relative mositure data and where the communities in this ecosystem stand.

```{r}
comm.mntd.mds <- monoMDS(comm.mntd.dist)
mds.fig <- ordiplot(comm.mntd.mds, type = "none")
points(mds.fig, "sites", pch = 19, col = "green", select = metadata$rel.moisture == "Fescue") 
points(mds.fig, "sites", pch = 19, col = "blue", select = metadata$rel.moisture == "Mixedgrass")
ordiellipse(comm.mntd.mds, metadata$rel.moisture, conf = 0.95, label = TRUE)

comm.mntd.traits.mds <- monoMDS(comm.mntd.traits.dist)
mds.fig <- ordiplot(comm.mntd.traits.mds, type = "none")
points(mds.fig, "sites", pch = 19, col = "green", select = metadata$rel.moisture == 
    "Fescue")
points(mds.fig, "sites", pch = 19, col = "blue", select = metadata$rel.moisture == 
    "Mixedgrass")
# add confidence ellipses around habitat types
ordiellipse(comm.mntd.traits.mds, metadata$rel.moisture, conf = 0.95, label = TRUE)


```

Testing for multivariate differences among groups

Adonis2 (the updated version of Adonis) is utilized for analyzing and comparing sum of squares results for multiple variables at once: in this case, we are looking at trait distance and relative moisture for community metadata. This is like ANOVA, but multivariate. 

The results here show something similar to the output of running ANOVA, including the p-value as well, demonstrating significance at 0.001. However, here there are also formula for each output, interpreted by what was placed in the parentheses for each execution of adonis2. The long list of numbers is the 999 permutations.

#Adonis 
```{r}

adonis2(comm.bc.dist ~ rel.moisture, data = metadata)
adonis2(comm.bc.dist ~ rel.moisture, data = metadata)
adonis2(comm.mntd.traits.dist ~ rel.moisture, data = metadata)

```

